---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## Installation

You can download the development version from GitHub with:

```{r installation, eval = FALSE}
if(!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("jmbadia/MS2ID", force=T)
```

## Introduction

This package annotates MS2 query spectra by using an in-house database. The creation of such database from (mostly publicly) available resources (HMDB, ChEBI, PubChem...) relies on the use of the [CompoundDb](https://github.com/EuracBiomedicalResearch/CompoundDb) package developed by J. Stanstrup and J. Rainer. 

For more information visit the package [website](https://jmbadia.github.io/MS2ID/articles/MS2ID.html) or [vignette.](https://github.com/jmbadia/MS2ID/tree/master/vignettes)

Discussions and suggestions are welcome: https://github.com/jmbadia/MS2ID/issues

