---
title: "Annotate query spectra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annot object: Working with results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
*This vignette describe in detail the function `annotate`. It is advised a previous reading of the MS2ID introduction.  

## Introduction  
`annotate()` is the MS2ID function that annotates MS/MS query spectra. That is, every query spectrum is compared with a reference library, and compounds with similar spectrum are listed.

Annotation function requires a reference library as MS2ID object, as described in the more general vignette Introduction to MS2ID vignette. 
```{r CompDbFormation, message=FALSE, cache=TRUE}
library(CompoundDb)
wrkDir <- tempdir()
## Locate the compounds file
MoNAsubset <- system.file("extdata/MoNAsubset.sdf.gz", package = "MS2ID")
cmps <- compound_tbl_sdf(MoNAsubset)
spctr <- msms_spectra_mona(MoNAsubset, collapsed = TRUE)
spctr$predicted <- FALSE
#configure metadata  
metad <- data.frame(
        name = c("source", "url", "source_version","source_date"),
        value = c("MoNA", "https://mona.fiehnlab.ucdavis.edu/downloads",
                  "v1", '07-09')
    )
#obtain compdb object
cmpdbDir <- file.path(wrkDir, "cmpdbDir")
if(dir.exists(cmpdbDir)){
  do.call(file.remove, list(list.files(cmpdbDir, full.names = TRUE)))
}else{
  dir.create(cmpdbDir)
}
db_file <- createCompDb(cmps, metadata = metad, msms_spectra = spctr,
                        path = cmpdbDir)
cmpdb <- CompDb(db_file)
library(MS2ID)
MS2IDdirectory <- createMS2ID(cmpdb = cmpdb, overwrite = TRUE,
                              path = wrkDir)
```

The following example shows basic usage of the function annotate taking advantage of tools developed in the Spectra package; please note that directly loading mzML files is also possible.  

```{r SimpleAnnotation, message=FALSE}
library(utils, Spectra, MS2ID)
queryFile <- system.file("extdata/Met1.zip", package = "MS2ID")
queryFolder <- file.path(wrkDir, "QRYspectra")
utils::unzip(queryFile, exdir = queryFolder)
querySpectra <- Spectra::Spectra(dir(queryFolder, full.names = TRUE))
querySpectra <- querySpectra %>%
  Spectra::filterMsLevel(2) %>% 
  Spectra::filterRt(c(100, 400))
refLibrary <- MS2ID(MS2IDdirectory)
annotResult <- annotate(QRYdata = querySpectra, MS2ID = refLibrary)
```
The annotate() function returns an Annot object.  

## Annot object
This object stores the annotation so we can:

* Browse graphically results with the MS2Igui() function.
* Export them to an xlsx file by using the export2xlsx() function
* Extract particular content by applying them *getters* (custom functions to *get* the info).
    + **hits()**: returns a cross-reference dataframe containing the hits along with their proposed adducts and common masses.
    + **qrySpectra()**: returns an Spectra object (see [Spectra](https://www.bioconductor.org/packages/release/bioc/html/Spectra.html) package) containing the query spectra with hits.
    + **refSpectra()**: returns an Spectra object with the successful reference spectra.
    + **refCompound()**: returns a dataframe containing (reference) compound metadata of successful reference spectra.
    + **infoAnnotation()**: returns the variables used on the annotation process.

## Consensus stuff  
Lorum ipsum.

## Subsetting the reference library
In annotation, subsetting the reference library not only reduces significantly the computing time (taking advantage of the MS2ID backend and the fragments index); it also prunes the result and cuts off non-sense hits. For example, the `cmnFrags = (m, n)` argument limits the reference spectra so that it is fulfilled that reference spectra and query spectra have at least m peaks in common among their top n most intense peaks,
In addition, the `cmnPrecMass` argument limits the reference spectra to those that have the same precursor mass as the query spectrum. And `cmnNeutralMass` will keep only those with a neutral mass that matches some of the plausible neutral masses of the query (considering the query precursor mass and all possible adducts).
```{r subsetting, eval=FALSE}
annotResult <- annotate(QRYdir = queryDir, MS2ID = refLibrary, 
                        cmnFrags = (3, 5),
                        cmnPrecMass = TRUE, cmnNeutralMass = TRUE, 
                        cmnPolarity = TRUE, nature = "experimental")
```

Other arguments subset the reference library according its nature or the polarization of the reference spectra (`predicted` and `cmnPolarity` respectively).

## Annotation with different metrics
As a default, annotate function use cosine similarity as a metric to compare two spectra; also, its default threshold value to beat in order to consider the comparison a hit is 0.8.  
The function also allows the simultaneous calculation of different metrics. In that case, a spectrum comparison is considered a hit when at least one of the metrics exceeds its threshold value. Note that *to beat* a threshold value has a different meaning depending on the metric. *Topsoe* and *squared_chord* metrics return a lower number when the spectra are more similar so, unlike the rest, a hit will occur when the returned value will be lower than its threshold. 
```{r diffMetrics, echo=TRUE, message=FALSE}
annotResult <- annotate(QRYdata = querySpectra, MS2ID = refLibrary,
                        metrics = c("fidelity", "cosine", "topsoe"),
                        metricsThresh = c(0.6, 0.8, 0.6))
head(MS2ID::hits(annotResult))
```

Moreover, the user can define its own metric by declaring a function as an argument, along with its threshold value; in the following example, foo function uses cosine+1 as a distance metric. 
```{r fooAnnotation, echo=TRUE, message=FALSE}
foo <- function(finalMatrix){
  vector1 <- finalMatrix[1,]
  vector2 <- finalMatrix[2,]
  CosplusOne <- 1+ suppressMessages(
    philentropy::distance(rbind(vector1, vector2), method = "cosine")
    )
  names(CosplusOne) <- "CosplusOne"
  return(CosplusOne)
}
annotResult <- annotate(QRYdata = querySpectra, MS2ID = refLibrary,
                        metrics = c("cosine"), metricsThresh = c(0.8),
                        metricFUN = foo, metricFUNThresh = 1.8)
head(MS2ID::hits(annotResult))
```

